Supplementary Note 1: Evaluation metrics

### 2018-08-24 manuscript/supplementary_notes/metrics/metrics.Rmd
update metrics

Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models. We therefore sought to use different comparison metrics, each serving a different purpose:

@@

The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$. To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$.

## $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("isomorphic", "latex")`$: Number of edge additions and removals to convert one topology to another

@@

@@

Although most metrics described above already assess some aspects directly relevant to the user (eg. whether the method is good at finding the right topology),

## $`r label_metric("featureimp_cor", "latex")`$: The similarity between features differentially expressed along the trajectory

>>>>

Given the multilayered complexity of a trajectory model, it is not trivial to compare the similarity of two trajectory models using only one metric. We therefore sought to use different comparison metrics, each serving a different purpose:

@@

## $`r label_metric("isomorphic", "latex")`$, $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$: Edit distance between two trajectory topologies

@@

Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone _or_ branch `r ref("fig", "clustering_scores_overview")`. These clusters of cells can then be compared using one of the many external cluster evaluation measures [@saelensComprehensiveEvaluationModule2018]. When selecting such a metric, we had two conditions:

- Because we allow some methods to filter cells in the trajectory, the metric should be able to handle "non-exhaustive assignment", where some cells are not assigned to any cluster.
- The metric should give each cluster equal weight, so that rare cell stages are equally important as large stages.

Based on these requirements, and on the analysis in [@saelensComprehensiveEvaluationModule2018], we chose the $\textrm{F1}$ score between the $\textrm{Recovery}$ and $\textrm{Relevance}$. If

```{r, results = "asis"}
add_fig(
  result_file("plot_clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
  "clustering_scores_overview",
  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ .")
)
```

The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$. To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$. TODO

@@

Although most metrics described above already assess some aspects directly relevant to the user, such as whether the method is good at finding the right topology, these metrics do not assess the quality of downstream analyses and hypotheses which can be generated from these models.

## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of differentially expressed features/genes along the trajectory

One of the main advantages of studying cellular dynamic processes using single-cell -omics data is that the
